Performance portability of an intermediate-complexity atmospheric research model in coarray Fortran

### 2017-10-03 body.tex
Committing edits from August 16.

\begin{figure}
   \vspace{-18pt}
   % Animate frames at 7 fps created via 'ffmpeg -i qvmovie.mp4 -r 7 frame%d.pngr', add control buttons,
   % and automatically play when the page is viewed:
   \vbox{\hspace{-24pt}
   %\animategraphics[width=1.1\columnwidth,controls,autoplay]{7}{figures/icar/frames/frame}{1}{30}
   \includegraphics[width=1.1\columnwidth]{figures/icar/frames/frame22.png}
   }
   \caption{An animation of atmospheric flow of water vapor (blues) and resulting precipitation (green to red) simulated by the \href{https://github.com/gutmann/icar}{\gls{icar}}. {\it For the animated version of this graphic, please access https://github.com/sourceryinstitute/coarray-icar-paw17/releases/download/1.0/main.pdf and use Adobe Acrobat Reader 7.0 or later.}
\label{figure:icar}}
\end{figure}

@@

\subsection{Objectives}
The goal of this work is to evaluate several alteranatives for compiling, linking, and executing the
same \gls{mini-app} \gls{caf} source code using the following technologies:
\begin{itemize}
  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{chapman2010introducing} communication layers,
  \item \gls{gcc} and Cray compilers, and
  \item Multi- and many-core processors.
\end{itemize}
Additionally we show performance differences between two one-sided data access patterns:
\begin{enumerate}
  \item Gets: an image that needs data proactively gets the data from memory managed by a providing
        image without the providing image's active involvement, and
  \item Puts: an image that is able to provide data proactively puts the data into memory managed by
        a receiving image without the receiving image's active involvement.
\end{enumerate}
Our results will inform our future decisions around the choice of compiler, communication
layer, platform, and data access patterns.

@@

\section{Discussion of Results}
Results with two different domain sizes (500 x 500 and 2000 x 2000 grid cells) are presented in figure~\ref{fig1-2} for simulations performed on Cheyenne.
These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using \gls{mpi} or OpenSHMEM for communication.
The differences between \gls{mpi} and OpenSHMEM are minor, with OpenSHMEM performing only slightly better.
However, the differences between the ``get'' and ``put'' operations are substantial.
For the 500x500 domain, ``puts'' scale well to 1800 cores, then scale poorly to \num{3600} cores;
however, ``gets'' only scale well to \num{504} cores, scale poorly to \num{1224}, an no further improvement comes from up to \num{3600} cores.
Note \gls{mpi} ``gets'' are not reported because a problem with the \gls{mpi} implementation makes them several orders of magnitude less efficient,
similarly, \gls{mpi} ``puts'' results are not reported beyond \num{1800} cores because a problem in the \gls{mpi} implementation caused memory allocations to become a bottleneck at higher process counts.
For the 2000x2000 domain, the same pattern is observed but scaling of ``puts'' remains strong up to 3600 cores;
scaling of ``gets'' continuous, albeit poorly, to \num{3600} cores.

Results across multiple compilers, machines, and architectures are presented in figure~\ref{fig3-4} (top).
In this case, only ``puts'' are tested, and only for the 2000x2000 domain.
The GNU compiler was used on Cheyenne (Xeons), and the Cray compiler was used on both Edison (Xeons) and Cori (KNL).
To aid interpretation, the scaling results are reported as a fraction of the ideal scaling for each machine in the top left.
These results show that the Cray compiler+system scales better than the gfortran+SGI system,
with 75\% efficiency at >10k cores, while on Cheyenne, only 55\% of ideal was achieved.
This also shows that the KNL system scales well out to large core counts (60\% ideal with \num{19200} cores),
but that the total runtimes are significantly slower then the equivalent runtimes on Xeons (top right).
The KNL performance may be improved in the future by implementing OpenMP threaded parallelism within a node.

To further test the scaling performance in the best case, ``puts'' on Cray Xeons, a limited set of tests were performed up to nearly \num{100000} cores (fig.~\ref{fig3-4} bottom).
This system scaled well to over \num{10000} cores, and continued scaling with nearly 50\% of single node efficiency obtained at \num{98304} cores.

>>>>

\begin{figure}
   \vspace{-18pt}
   % Animate frames at 7 fps created via 'ffmpeg -i qvmovie.mp4 -r 7 frame%d.pngr', add control buttons,
   % and automatically play when the page is viewed:
   \vbox{\hspace{-24pt}
   %\animategraphics[width=\columnwidth,controls,autoplay]{7}{figures/icar/frames/frame}{1}{30}
   \includegraphics[width=1.1\columnwidth]{figures/icar/frames/frame22.png}
   }
   \caption{An animation of atmospheric flow of water vapor (blues) and resulting precipitation (green to red) simulated by the \href{https://github.com/gutmann/icar}{\gls{icar}}. {\it For the animated version of this graphic, please access https://github.com/sourceryinstitute/coarray-icar-paw17/releases/download/1.0/main.pdf and use Adobe Acrobat Reader 7.0 or later.}
\label{figure:icar}}
\end{figure}

@@

\subsection{Objectives}
The goal of this work is to evaluate several alteranatives for compiling, linking, and executing the
same \gls{mini-app} \gls{caf} source code using the following technologies:
\begin{itemize}
  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{chapman2010introducing} communication layers,
  \item \gls{gcc} and Cray compilers, and
  \item Multi- and many-core processors.
\end{itemize}
Additionally we show performance differences between two one-sided data access patterns:
\begin{enumerate}
  \item Gets: an image that needs data proactively gets the data from memory managed by a providing
        image without the providing image's active involvement, and
  \item Puts: an image that is able to provide data proactively puts the data into memory managed by
        a receiving image without the receiving image's active involvement.
\end{enumerate}
Fortran semantics necessitates that gets block.  Puts are non-blocking.

Our results will inform our future decisions around the choice of compiler, communication
layer, platform, and data access patterns.

@@

\section{Discussion of Results}
Figure~\ref{fig1-2} shows results with two domain sizes (500 x 500 and 2000 x 2000 grid cells) running on Cheyenne.
These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using \gls{mpi} or OpenSHMEM.
The differences between \gls{mpi} and OpenSHMEM are minor, with OpenSHMEM performing only slightly better.
However, the differences between the ``get'' and ``put'' operations are substantial.
For the 500x500 domain, ``puts'' scale well to 1800 cores, then scale poorly to \num{3600} cores;
however, ``gets'' only scale well to \num{504} cores, scale poorly to \num{1224}, and no further improvement comes from up to \num{3600} cores.

\gls{mpi} ``gets'' are not reported because a problem with the \gls{mpi} implementation makes them several orders of magnitude less efficient.
Similarly, \gls{mpi} ``put'' results are not reported beyond \num{1800} cores because a problem in the \gls{mpi} implementation caused memory
allocations to become a bottleneck at higher process counts.
For the 2000x2000 domain, the same pattern is observed but scaling of ``puts'' remains strong up to 3600 cores; whereas
scaling of ``gets'' continues, albeit poorly, to \num{3600} cores.

Figure~\ref{fig3-4} (top) depicts results across multiple compilers, machines, and architectures.
In this case, only ``puts'' are tested, and only for the 2000x2000 domain.
The GNU compiler was used on Cheyenne (Xeons). The Cray compiler was used on Edison (Xeons) and Cori (KNL).
To aid interpretation, the scaling results are reported as a fraction of the ideal scaling for each machine in the top left.
These results show that the Cray compiler+system scales better than the gfortran+SGI system,
with 75\% efficiency at >10k cores, while on Cheyenne, only 55\% of ideal was achieved.
This also shows that the KNL system scales well out to large core counts (60\% ideal with \num{19200} cores),
but that the total runtimes are significantly slower than the equivalent runtimes on Xeons (top right).
The KNL performance might be improved in the future by implementing OpenMP threaded parallelism within a node.

To further test the scaling performance in the best case, ``puts'' on Cray Xeons, a limited set of tests were performed up to nearly \num{100000} cores (fig.~\ref{fig3-4} bottom).
This system scaled well to over \num{10000} cores, and continued scaling with nearly 50\% of the single-node efficiency at \num{98304} cores.
